\documentclass[12pt,a4paper]{article}

% --- PACKAGES ---
\usepackage{amsmath}        % For advanced math environments (align, gather, etc.)
\usepackage{amssymb}        % For mathematical symbols (e.g., \mathbb, \mathfrak)
\begin{document}




\vspace{0.5cm}



\section{Problems}

\begin{enumerate}
    \item Let $D_4$ be the dihedral group of symmetries of a square.
    \begin{enumerate}
        \item How many elements does $D_4$ have?
        \item How many of these are rotations?
        \item How many are reflections?
    \end{enumerate}

    \item Find all subgroups of $D_4$. Which of them are cyclic?

    \item Let $S_3$ be the symmetric group on 3 elements. Let $\sigma = (1\ 2)$ and $\tau = (1\ 2\ 3)$.
    \begin{enumerate}
        \item Compute $\sigma\tau$ and $\tau\sigma$ and compare.
        \item List all subgroups of $S_3$. Identify which are cyclic and which are not.
    \end{enumerate}

    \item Prove that  set of all rotations in $D_n$ form a subgroup of $D_n$.
    \item Let $G$ be a group and $H_1$ and $H_2$ be two subgroups of $G$.
    \begin{enumerate}
    \item Prove that $H_1\cap H_2$ is a subgroup and also  $\bigcap H_\alpha$, $\alpha \in I$ where $I$ is an indexing set
    \item what about $H_1 \cup H_2$? 
\end{enumerate}
\item Verify that $GL_n(\mathbb{R})$ satisfies all four group axioms under matrix multiplication. Define the special linear group $SL_2(\mathbb{R})$ as the set of $2 \times 2$ real matrices with determinant 1.
\begin{enumerate}
    
  \item[(a)] Show that $SL_2(\mathbb{R})$ is a subgroup of $GL_2(\mathbb{R})$.
  \item[(b)] Verify whether the matrix $B = \begin{bmatrix} 2 & 1 \\ 1 & 1 \end{bmatrix}$ belongs to $SL_2(\mathbb{R})$.
  
\end{enumerate} 
\item Let  $\mathbb{Z}$ be the group of integers and let $k,l \in \mathbb{Z}$ .What is the subgroup generated by $k$ and $l$? % Some space at the end
\item Is  it possible that the subgroup generated by $2$ and $3$ in $\mathbb{Q}$ under addition is cyclic?
\begin{enumerate}
    \item What about the subgroup generated by any two elements in $\mathbb{Q}$?
    \item What about subgroup generated by $2$ and $3$ in $\mathbb{Q}-\{0\}$ under multiplication?  
\end{enumerate}
\item Define $det : GL_n(\mathbb R) \to \mathbb R$ as the determinant map. Show that this is a group homomorphism-

    \begin{enumerate}
        \item What is the kernel of this homomorphism?
        \item What is its image?
        \end{enumerate}
\item Let $H=< P_\sigma  ; \sigma \in S_n>$ is a subgroup of $S_n$. Let $\phi$ is a map from $S_n$ to $GL_n(\mathbb{R})$ defined as $\phi(\sigma)=P_\sigma$. Prove that $\phi$ is one to one homomorphism.
%\item Let $n>m$  for $n,m \in \mathbb{N}$ what are the possible homomorphisms from,
%\begin{enumerate}
 %   \item $S_n$ to $S_m$
%    \item$S_m$ to $S_n$
%\end{enumerate}
\item Verify that $GL_n(\mathbb{R})$ satisfies all four group axioms under matrix multiplication. Define the special linear group $SL_2(\mathbb{R})$ as the set of $2 \times 2$ real matrices with determinant 1.
\begin{enumerate}
    
  \item Show that $SL_2(\mathbb{R})$ is a subgroup of $GL_2(\mathbb{R})$.
  \item Verify whether the matrix $B = \begin{bmatrix} 2 & 1 \\ 1 & 1 \end{bmatrix}$ belongs to $SL_2(\mathbb{R})$.
  
\end{enumerate} 
\end{enumerate}


\end{document}